---
title: "STAT 5605 Homework 4"
subtitle: "February 20, 2026"
author: "Jack Bienvenue"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    latex_engine: pdflatex
    keep_tex: true 
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{1em} 
fontsize: 12pt
---

```{r setup, include=FALSE}
invisible(lapply(c("ggplot2", "dplyr", "tidyr", "car", "broom"), 
  function(pkg) suppressPackageStartupMessages(library(pkg, 
                                                character.only = TRUE))))
```

# Problem 1

The dataset hw4.csv contains the following variables:

\begin{verbatim}
    Y: MPG
    X1: DISPLACEMENT
    X2: HORSEPOWER
    X3: WEIGHT
\end{verbatim}

```{r}
carmpg <- read.csv("./hw4.csv")
str(carmpg)
mod <- lm(y~x1+x3+x2, data=carmpg)
summary(mod)
anova(mod)
SSR=sum((fitted(mod)-mean(carmpg$y))^2)
SSR
SSTO=sum((carmpg$y-mean(carmpg$y))^2)
SSTO
Anova(mod, type = "III")
avPlots(mod)
```

\newpage

## (a)  

Obtain the analysis of variance table that decomposes the regression sum of squares into extra sums of squares associated with X1; with X3 given X1; and with X2 given X1 and X3.

```{r}
# Sequential ANOVA (Type I) will help with solving this problem:
anova(lm(y ~ x1 + x3 + x2, data = carmpg)) # type 1 for the order below
```

\begin{bluebox}
The problem defines a regression model for vehicle miles per gallon as:

$$
Y_i = \beta_0 + X_{i1}\beta_1 + X_{i2}\beta_2 + X_{i3}\beta_3 +\epsilon_i,
$$

Let's work on building the ANOVA tables with the desired composition. We can do this using sequential sums of squares where we decompose our sums of squares into $SSTO = SSR_{X_1} + SSR_{(X_3 | X_1)} + SSR_{(X_2|X_1,X_3)}+SSE$. The table becomes:

\textbf{Extra SS associated with $X_1$}:

\begin{tabular}{lccccc}
    \toprule
    Source & df & SS & MS & F & $p$ \\
    \midrule
    $X_1$ & 1 & 15440.2 & 15440.2 &  858.27 &  < 2.2e-16 \\
    $(X_3 | X_1)$ & 1 & 1208.5 & 1208.5 & 67.178 & 3.654e-1 \\
    $(X_2 | X_1, X_3)$ & 1 & 190.3  & 190.3 & 10.577 & 0.001245 \\
    Error      & 388 & 6980 & 18.0 &  &  \\
    \midrule
    Total               & 391 & 23819 &  &  &  \\
    \bottomrule
\end{tabular}

\end{bluebox}

## (b)  

Test whether X2 can be dropped from the regression model given that X1 and X3 are retained. Use the $F$ test statistic and $\alpha=0.05$. State the alternatives, decision rule, and conclusion. What is the P-value of the test?

```{r}
# Calculate critical value of F for this problem:
q1_crit_F <- qf(0.95, df1 = 1, df2 = 388)
# Here, numerator df is # of restrictions,
# denominator df is n-p-1 = 392-3-1
cat("Critical F value:", q1_crit_F)
```

\begin{bluebox}
We can test whether $X_2$, the engine horsepower, can be dropped from the regression model on the condition that $X_1$, the engine displacement (cc), and $X_3$, the vehicle weight, with the hypotheses:

$$
H_0: \beta_2 = 0, \text{ and }
$$

$$
H_1: \beta_2 \neq 0,
$$

Performing a partial F test. Our test statistics will be:

$$
F_{\text{test}} = \frac{(SSR_{X_2| X_1, X_3})/1}{MSE_{X_1,X_2,X_3}} =  \frac{190.3}{18} = 10.57
$$

As the question asks, we will state our test results using both the methods of the test statistic and p-value. 

We will check whether out test statistic exceeds its critical value of $3.866$ and we will use a significance threshold of $\alpha = 0.05$ to check whether our attained p-value falls below this threshold. 

Upon performing our test, we find that our test statistic exceeds the critical value ($10.57 > 3.866$), therefore we will reject the null hypothesis which stated that the slope associated with $X_2$ is zero when $X_1$ and $X_3$ are already involved in the model. 

The p-value for this test is given by the R output as 0.001245, which is below our significance threshold, verifying that we can reject the null hypothesis. 

\end{bluebox}


## (c)  

Does SSR(X1)$+$SSR(X2|X1) equal SSR(X2)$+$SSR(X1|X2) here? Must this always be the case?

```{r}
# x1, x2 order:
anova(lm(y~x1+x2, data = carmpg))
# x2, x1 order:
anova(lm(y~x2+x1, data = carmpg))
```



\begin{bluebox}
Let's find $SSR(X_1)+SSR(X_2|X_1)$ and $SSR(X_2)+SSR(X_1|X_2)$ to judge if they are equal.

$$
SSR(X1)+SSR(X2|X1) = 15440.2 + 383.5 = 15823.7, \text{ and }
$$

$$
SSR(X2)+SSR(X1|X2) = 14433.1 + 1390.6 = 15823.7,
$$

Therefore, we find that $SSR(X_1)+SSR(X_2|X_1)$ and $SSR(X_2)+SSR(X_1|X_2)$. 

This will always be the case, because:

$$ 
SSR(X_2|X_1) = SSR(X_1,X_2) - SSR(X_1) \implies SSR(X_2|X_1) + SSR(X_1) = SSR(X_1,X_2) $$

This is symmetric with $SSR(X_1|X_2)$:

$$ SSR(X_1|X_2) = SSR(X_2,X_1) - SSR(X_2) \implies SSR(X_1|X_2) + SSR(X_1) = SSR(X_2,X_1),$$

Since $SSR(X_2, X_1) = SSR(X_1,X_2),$ the two expressions $SSR(X_1)+SSR(X_2|X_1)$ and $SSR(X_2)+SSR(X_1|X_2)$ are always equal.
\end{bluebox}

## (d) 

Calculate $R^2(X_1)$, $R^2(X_2)$, $R^2(X_1, X_2)$, $R^2_{Y1|2}$, $R^2_{Y2|1}$, $R^2_{Y2|1,3}$, and $R^2$. Explain what each coefficient measures and interpret your results.

\begin{bluebox}
Let's carefully calculate all of these variations of the coefficient of determination:

$$
R^2(X_1) = \frac{SSR(X_1)}{SSTO} = \frac{15440.2}{23819} = 0.648,
$$

This coefficient represents the variation in $Y$ that can be explained by variations in $X_1$ when we regress $Y$ on $X_1$. Here, we can explain $64.8\%$ of the variation in vehicle gas mileage (mpg) through variations in engine displacement. 

$$
R^2(X_2) = \frac{SSR(X_2)}{SSTO} = \frac{14433.1}{23819} = 0.606,
$$
This coefficient represents the variation in $Y$ that can be explained by variations in $X_1$ when we regress $Y$ on $X_1$. Here, we can explain $60.6\%$ of the variation in vehicle mpg through variations in vehicle horsepower. 

$$
R^2(X_1, X_2) = \frac{SSR(X_1, X_2)}{SSTO} = \frac{15823.7}{23819} = 0.664,
$$
This coefficient represents the variation in $Y$ that can be explained by variations in $X_1$ and $X_2$ when we regress $Y$ on both $X_1$ and $X_2$. So, we can explain $66.4\%$ of the variation in vehicle mpg on variations in engine displacement and vehicle horsepower. 

$$
R^2_{Y1|2} = \frac{SSR(X_1 | X_2)}{SSE(X_2)} = \frac{1390.6}{23819 - 14433.1} = 0.148,
$$
This coefficient represents the additional variation in $Y$ explained by adding $X_1$ to a model already containing $X_2$. So, adding vehicle engine displacement to our model helps explain an additional $14.8\%$ of the variance in mpg compared to a model with only information about the variation in engine horsepower.

$$
R^2_{Y2|1} = \frac{SSR(X_2|X_1)}{SSE(X_1)} = \frac{383.5}{23819 - 15440.2} = 0.048,
$$

This coefficient represents the additional variation in $Y$ explained by adding $X_2$ to a model already containing $X_1$. So, adding vehicle horsepower to our model helps explain an additional $4.8\%$ of the variance in mpg compared to a model with only information about the variation in engine displacement.

$$
R^2_{Y2|1,3} = \frac{SSR(X_2 | X_1, X_3 )}{SSE(X_1, X_3)} = \frac{190.3}{23819 - 15440.2-1208.5} = 0.027,
$$

This coefficient represents the additional variation in $Y$ explained by adding $X_2$ to a model already containing $X_1$ and $X_3$. So, adding vehicle horsepower to our model helps explain an additional $2.7\%$ of the variance in mpg compared to a model with only information about the variation in engine displacement and vehicle weight.

$$
R^2 = \frac{SSR}{SSTO} = \frac{16839}{23819} = 0.707,
$$

This coefficient represents the variation in $Y$ that can be explained by variations in $X_1, X_2, \text{ and }, X_3$, the full model. Here, $70.7\%$ of the variation in mpg can be explained by variations in engine displacement (cc), engine horsepower, and vehicle weight. 

\end{bluebox}

## (e) 

Fit a regression model $Y_i=\beta_0+\beta_1X_{i1}+\beta_2X_{i2}+\varepsilon_i$ to the data using X1 and X2 only.

```{r}
q1e_model <- lm(y ~ x1 + x2,  data=carmpg)
summary(q1e_model)
```

## (f)  

For the model in part (e), prepare an added-variable plot for each of the predictor variables X1 and X2.

```{r}
# Added-variable plot for X1:
## Residuals of Y ~ X2
res_y_x2 <- resid(lm(y ~ x2, data = carmpg))
## Residuals of X1 ~ X2
res_x1_x2 <- resid(lm(x1 ~ x2, data = carmpg))

av_x1 <- data.frame( res_x1 = res_x1_x2, res_y  = res_y_x2)

ggplot(av_x1, aes(x = res_x1, y = res_y)) +
  geom_point(color = "forestgreen", size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "#400", linewidth = 1) +
  labs(title = "Added-Variable Plot for X1",
    subtitle = "Effect of X1 on Y Adjusting for X2",
    x = "Residuals of X1 | X2",
    y = "Residuals of Y | X2") + theme_minimal(base_size = 14)

# Added-variable plot for X2:
## Residuals of Y ~ X1
res_y_x1 <- resid(lm(y ~ x1, data = carmpg))
## Residuals of X2 ~ X1
res_x2_x1 <- resid(lm(x2 ~ x1, data = carmpg))

av_x2 <- data.frame( res_x2 = res_x2_x1, res_y  = res_y_x1)

ggplot(av_x2, aes(x = res_x2, y = res_y)) +
  geom_point(color = "forestgreen", size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "#400", linewidth = 1) +
  labs(
    title = "Added-Variable Plot for X2",
    subtitle = "Effect of X2 on Y Adjusting for X1",
    x = "Residuals of X2 | X1",
    y = "Residuals of Y | X1"
  ) +
  theme_minimal(base_size = 14)
```

## (g)  

Do your plots in part (f) suggest that the regression relationships in the fitted regression function in part (e) are inappropriate for any of the predictor variables? Explain.

\begin{bluebox}
The plots from part (f) do not suggest that the regression relationships in the regression function $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2X_{i2} + \epsilon_i$ are inappropriate. Our added-variable plots are point clouds that appear to have relatively constant variance and no obvious systematic departures in linearity.
\end{bluebox}

## (h)  

Suppose we want to see if it is the sum of DISPLACEMENT, HORSEPOWER, and WEIGHT that is really affecting the MPG. Perform an appropriate test to answer this question.

```{r}
# Perform appropriate F-test:
## Build proposed model:
carmpg$x4 <- carmpg$x1 + carmpg$x2 + carmpg$x3
q1_h_model <- lm(y ~ x4, data = carmpg)

anova(q1_h_model, mod)
```

\begin{bluebox}
We can evaluate the capability of a random variable $X_4 = X_1+X_2+X_3$ by using this new random variable to regress $y$ (MPG) upon, and compare the results of our new regression to the modeling containing all predictors separately.

Since the sum $X_4 = X_1+X_2+X_3$ would indicate that all three variables would have equal weighting, we can structure our test in the following way:

$$
H_0: \beta_1 = \beta_2 = \beta_3, \text{ versus}
$$

$$
H_1: \text{At least two } \beta_i \; (i=1,2,3) \text{ differ}.
$$

To run our test, let's prespecify a significance level of $\alpha = 0.05$. Our test results using a partial F test give a p-value of $0.009059$, far below our threshold. Therefore, we can reject the null hypothesis which stated that the coefficients for the three predictors are identical. Therefore, we have evidence to suggest that the individual effects of the predictors in combination have a better explanatory power over the variation in mpg than the effect of the sum of the predictors.
\end{bluebox}






\newpage
# Problem 2

For the data $(X_1,X_2,Y)=(-1,-1,7.2)$, $(-1,0,8.1)$, $(0,0,9.8)$, $(1,0,12.3)$, $(1,1,12.9)$, consider the model:   $E(Y)=\beta_0 + \beta_1 X_1 + \beta_2 X_2$.

## (a)  

```{r}
# Let's make our data into a data frame for our calculations:
q2_x1 <- c(-1, -1, 0, 1, 1)
q2_x2 <- c(-1, 0, 0, 0, 1)
q2_y <- c(7.2, 8.1, 9.8, 12.3, 12.9)

q2_df <- data.frame(
  x1 = q2_x1,
  x2 = q2_x2,
  y = q2_y
)

print(q2_df)

## Now run the regressions to get all necessary sums of squares for coefficient 
## of determination calculations:

q2a_lm_x1 <- lm(y ~ x1, data=q2_df)
anova(q2a_lm_x1)

q2a_lm_x1x2 <- lm (y ~ x1 + x2, data = q2_df)
summary(q2a_lm_x1x2)
anova(q2a_lm_x1x2)
```

Let $R^2(X_1)$ and $R^2(X_1,X_2)$ denote the coefficients of determination ($R^2$'s) in the regressions of $Y$ on $X_1$ and $Y$ on $X_1$ and $X_2,$ respectively.  Compute $R^2(X_1)$, $R^2_{Y2|1}$, and $R^2(X_1,X_2)$.

\begin{bluebox}
Let's compute each of the coefficients of determination:

$$
R^2(X_1) = \frac{SSR(X_1)}{SSTO} = \frac{24.5025}{25.172} = 0.97340,
$$

$$
R^2_{Y 2|1} = \frac{SSR(X_2|X_1)}{SSE(X_1)} = \frac{0.5625}{0.6695} = 0.84018, \text{ and }
$$

$$
R^2(X_1, X_2) = \frac{SSR(X_1, X_2)}{SSTO} = \frac{24.5025+0.5625}{24.5025+0.5625+0.1070} = 0.99575.
$$
\end{bluebox}

## (b)  

Derive the relationship between $R^2(X_1,X_2)$ and $(R^2(X_1), R^2_{Y2|1})$.

\begin{bluebox}
The relationship between $R^2(X_1,X_2)$ and $(R^2(X_1), R^2_{Y2|1})$ can be derived by starting with a look at the expressions for each coefficient of determination. 

$$
R^2(X_1) = \frac{SSR(X_1)}{SSTO}; \quad R^2_{Y 2|1} = \frac{SSR(X_2|X_1)}{SSE(X_1)} ; \quad \text{ and } R^2_{X_1, X_2} = \frac{SSR(X_1, X_2)}{SSTO}
$$

To express $R^2(X_1,X_2)$ in terms of ($R^2(X_1)$, $R^2(X2|X_1)$), we have to recognize some helpful relationships for coefficients of determination. First, $R^2(X_1) = 1 - \frac{SSE(X_1)}{SSTO} \implies SSE(X_1) = SSTO(1-R^2(X_1))$. Next, we will want to come up with an expression for $SSR(X_2|X_1)$, $SSR(X_2|X_1) = SSR(X_2, X_1) - SSR(X_1)$ (from $SSR(X_1, X_2) = SSR(X_2 | X_1) + SSE(X_1)$. Using these expressions,

$$
R^2_{(X_1, X_2)} = \frac{SSR(X_1, X_2)}{SSTO} = \frac{SSR(X_2|X_1) +SSE(X_1)}{SSTO} = \frac{SSR(X_2|X_1)}{SSTO} + R^2(X_1),
$$
We don't have a particular coefficient of determination that lines up with $\frac{SSR(X_2|X_1)}{SSTO}$, but we can decompose this term by recognizing $R^2_{Y2|1} = \frac{SSR(X_2|X_1)}{SSE(X_1)},$ or $SSE(X_1) \times R^2_{Y2|1} = SSR(X_2|X_1).$ Substituting this in:

$$
R^2_{(X_1, X_2)} = \frac{SSE(X_1) \times R^2_{Y2|1}}{SSTO} + R^2(X_1) = R^2_{Y2|1}(1-R^2(X_1)) + R^2(X_1)
$$

So, we can actually express $R^2_{(X_1, X_2)}$ directly in terms of $\left(R^2_{Y2|1},R^2(X_1)\right)$.
\end{bluebox}

## (c)  

Empirically verify the relationship obtained in (ii) using the data given in this problem.

\begin{bluebox}
Let's verify that the relationship from part (b) holds for this particular example by calculating $R^2(X_1,X_2)$ using the derived relationship for this data set:

$$
R^2(X_1,X_2) = R^2_{Y2|1}(1-R^2(X_1)) + R^2(X_1) = 0.84018(1-0.97340) +0.97340 = 0.99575,
$$
Which is precisely the same result as when we calculated the value in Part (a).

\end{bluebox}

## (d)  

Compute $r_{YX_2\cdot X_1}$.  What is the relationship between $R^2_{Y2|1}$ and $r_{YX_2\cdot X_1}$?

```{r}
# Compute r_{YX_2\cdot X_1}$:
## Find residual terms:
res_Y <- resid(lm(y ~ x1, data = q2_df))
res_X2 <- resid(lm(x2 ~ x1, data = q2_df))

## Calculate correlation:
q2_d_cor <- cor(res_Y, res_X2)
cat(q2_d_cor)
```

\begin{bluebox}
Let's compute $r_{YX_2\cdot X_1}$:

$$
r_{YX_2\cdot X_1} = Corr(\hat{\epsilon}_{Y|X1},\hat{\epsilon}_{X_2|X_1}),
$$
Which we can compute in R to find:

$$
r_{YX_2\cdot X_1} = 0.9166
$$ 

So, these two quantities are related by the relationship $R^2_{Y2|1}$ = $r^2_{YX_2\cdot X_1},$ as we can see $R^2_{Y2|1} = 0.84018 = (0.9166)^2 = r^2_{YX_2\cdot X_1}$.

\end{bluebox}





\newpage
# Problem 3

The following regression model is being considered in a market research study:
$$
    Y_i=\beta_0+\beta_1X_{i1}+\beta_2X_{i2}+\beta_3X_{i1}^2+\varepsilon_i.
$$

Suppose that an R data frame, ``data", contains variables y, x1, and x2.
The following R code is used to produce the full model:

\begin{verbatim}
dat$x11=dat$x1^2
mod <- lm(y~x1+x2+x11, data=dat)
\end{verbatim}

State the reduced models and write the corresponding R codes for testing whether or not:

## (a) $\beta_1=\beta_3=0$
\begin{verbatim}
mod <- lm(y~x2, data=dat)
\end{verbatim}

```{r}
# Test for beta1=beta3=0:

```

\begin{bluebox}
To test whether $\beta_1 = \beta_3 = 0$, we must have reduced models MODELS!!!!!!!!

We find ???????
\end{bluebox}

## (b) $\beta_0=0$
\begin{verbatim}
dat$x11=dat$x1^2
mod <- lm(y ~ 0+x1+x2+x11, data=dat)
\end{verbatim}

```{r}
# Test for beta0 = 0
```

\begin{bluebox}
To test whether $\beta_0 = 0$, we must have reduced models MODELS!!!!!!!!

We find ???????
\end{bluebox}

## (c)  $\beta_3=5$
\begin{verbatim}
dat$x11=dat$x1^2
dat$x115=5*dat$x1^2
mod <- lm(y ~ x1+x2+offset(x115), data=dat)
\end{verbatim}

```{r}
# To test whether beta3 = 5,

``` 

\begin{bluebox}
To test whether $\beta_3 = 5$, we must have reduced models MODELS!!!!!!!

We find ????????
\end{bluebox}

## (d)  $\beta_0=10$

```{r}
# Test for \beta_0 = 10
```

\begin{bluebox}
To test whether $\beta_0 = 10$, we must have reduced models MODELS!!!!!!!

We find ????????
\end{bluebox}

## (e) $\beta_1=\beta_2$

```{r}
# Test for beta1 = beta2

```

\begin{bluebox}
To test whether $\beta_1 = \beta_2$, we must have reduced models MODELS!!!!!!!

We find ????????
\end{bluebox}

## (f) $\beta_1=\beta_3$

```{r}
# Test for beta1 = beta3

```

\begin{bluebox}
To test whether $\beta_1 = \beta_3$, we must have reduced models MODELS!!!!!!!

We find ????????
\end{bluebox}











\newpage
# Problem 4

Determine whether the following statements are true. Give a brief justification of your answer.

## (a) 

Suppose that we fit the model $Y=\beta_0 + \beta_1 X+\epsilon$ to a dataset $\{ (Y_i, X_i), \; i=1,2,\dots, n\}$ with $S_{YY}=\sum^n_{i=1}(Y_i-\bar{Y})^2 >0$, where $\bar{Y}=\frac{1}{n}\sum^n_{i=1} Y_i$. Then, all the residuals must be zero when $R^2=1$.

\begin{bluebox}
\textbf{TRUE}: 

We can define the coefficient of determination $R^2$ as $R^2 = 1-\frac{\sum(Y_i - \hat{Y})^2}{\sum(Y_i - \bar{Y})^2}$, so for this rhs expression to equal 1 when $R^2 = 1$, the residual sum of squares $\sum(Y_i - \hat{Y})^2$ must be zero. Since the residual sum of squares is a sum of non-negative values, the sum will only be equal if every summed element is zero. Here, each summed element is the squared residual, so the underlying residual must be zero. Therefore, all residuals must be 0 when $R^2=1$ given the total sum of squares is positive.

\end{bluebox}

## (b)  

When we fit the model $Y=\beta_0 +\epsilon$ (i.e., no $X$'s) to a set of data, $R^2=0$, always.

\begin{bluebox}
\textbf{TRUE}:

The coefficient of determination is formulated as $R^2 = 1 - \frac{\sum (y_i - \hat{y}_i)^2}{\sum (y_i - \bar{y})^2}$. For the mean-only model, the fitted values $\hat{y}_i$ are the same as the mean values $\bar{y}$. So, the numerator and denominator of the $\frac{SSE}{SSTO}$ fraction become equivalent, and so $R^2$ will always be $0$ for the mean model.

\end{bluebox}

## (c)  

The model $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_2 \log (X_1) + \epsilon$, where $X_1>0$, $E(\epsilon)=0$, and $\mbox{Var}(\epsilon)=\sigma^2>0$, is a linear model.

\begin{bluebox}
\textbf{TRUE}:
This model is linear in all its coefficients $\bm{\beta} = (\beta_0, \beta_1, \beta_2, \beta_3)^\top$, and therefore is a linear model for $Y$.
\end{bluebox}










\newpage
# Problem (5) 

An endocrinologist was interested in exploring the relationship between the level of a steroid (Y) and age (X) in healthy female subjects whose ages ranged from 8 to 25 years. She collected a sample of 27 healthy females in this age range. The data are given in ``Steroid$_{-}$level.txt''

```{r}
steroid <- read.table("./steroid_level.txt", header=TRUE)
str(steroid)
nrow(steroid) 
steroid$x2=steroid$x^2
PR5 <- lm(y~x+x2,data=steroid)
summary(PR5)
new_x_value <- data.frame(x = 15, x2=225)
predicted_value2 <- predict(object = PR5, newdata = new_x_value, interval="prediction", level=0.99)
print(predicted_value2)
```

## (a)  

Fit regression model $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon$. Plot the fitted regression function and the data. Does the quadratic regression function appear to be a good fit here? Find $R^2$.

```{r, fig.width = 8, fig.height = 5, fig.cap = "F Steroid Level Regression"}
# Fit regression model:

# Plot fitted regression fxn with data:


# Calculate R^2
```
  
\begin{bluebox}
The regression function does (not?) appear to be a good fit here.

We still \emph{can} calculate $R^2$ for this example, but its interpretation is no longer valid as our model is no longer linear in $X_i$'s. The $R^2$ value is $.$.
\end{bluebox}

## (b) 

```{r}
# Perform hypothesis test

```

Test whether or not there is a regression relation; use $\alpha=0.01$. State the alternatives, decision rule, and conclusion. What is the P-value of the test?
  
\begin{bluebox}
To test for whether there is a regression relation between $Y$ and $X$ \& $X^2$, we can use the hypotheses:

$$
H_0: , \text{ and }
$$

$$
H_1: , \text{ and }
$$

After performing our test using a prespecified significance level of $\alpha = 0.01$, we find...

Therefore, we have evidence to suggest that there is (NOT) a regression relation between $Y$ and $X$ \& $X^2$.
\end{bluebox}

## (c) 

Predict the steroid levels of females aged 15 using a 99 percent prediction interval. Interpret your interval.

```{r}
# Calculate prediction interval (99%)

```

\begin{bluebox}
We can use a $99\%$ prediction interval to predict the steroid levels in $15$ year old girls, which ends up being $()$. This interval suggests that ...
\end{bluebox}

## (d)  

Test whether the quadratic term can be dropped from the model; use $\alpha=0.01$. State the alternatives, decision rule, and conclusion.

```{r}
# Perform test
```

\begin{bluebox}
We can test whether the quadratic term can be dropped from the model using hypotheses:

$$
H_0: \text{ and }
$$

$$
H_A: ,
$$

Which result in a p-value of $,$ which we compare to our prespecified significance level $\alpha = 0.01$ to find that ...

Therefore, we can conclude ...
\end{bluebox}

## (e)  

Obtain the residuals and plot them against the fitted values and against $x$ on separate graphs. Also prepare a normal probability plot. What do your plots show?

```{r, fig.cap ="a"}
# Plot of residuals vs fitted values

```


```{r, fig.cap ="b"}
# Plot of residuals vs x

```

```{r, fig.cap ="c"}
# Normal probability plot
```

\begin{bluebox}
Figure . shows...

Figure . shows...

Figure . shows...
\end{bluebox}

## (f)  

Test formally for lack of fit. Control the risk of a Type I error at .01. State the alternatives, decision rule, and conclusion. What assumptions did you make implicitly in this test?

```{r}

```

\begin{bluebox}
Now we can test for a lack of fit with a significance level $\alpha = 0.01$ and hypotheses:

$$
H_0: \text{ and }
$$

$$
H_1: 
$$

Upon running our test, we find...

Our assumptions for this test include...

\end{bluebox}











\newpage
# Problem 6

For predicting the number of active physicians (Y) in a county, it has been decided to include total population (X1) and total personal income (X2) as predictor variables. The question now is whether an additional predictor variable among land area (X3), percent of population 65 or older (X4), number of hospital beds (X5), and total serious crimes (X6) would be helpful in the model and, if so, which variable would be most helpful. Use the following R outputs to answer the following questions.

\begin{verbatim}
> anova(lm(Y ~ X1 + X2 + X3 + X4 + X5 + X6, data=dat))
Analysis of Variance Table
Response: Y
           Df     Sum Sq    Mean Sq   F value    Pr(>F)
X1          1 1243181164 1243181164 8851.8032 < 2.2e-16 ***
X2          1   22058054   22058054  157.0596 < 2.2e-16 ***
X3          1    4063370    4063370   28.9323 1.229e-07 ***
X4          1     608535     608535    4.3329   0.03797 *
X5          1   75286073   75286073  536.0582 < 2.2e-16 ***
X6          1     196924     196924    1.4022   0.23701
Residuals 433   60812180     140444
\end{verbatim}


\begin{verbatim}
> anova(lm(Y ~ X1 + X2 + X4 + X5 + X6 + X3, data=dat))
Analysis of Variance Table
Response: Y
           Df     Sum Sq    Mean Sq   F value  Pr(>F)
X1          1 1243181164 1243181164 8851.8032 < 2e-16 ***
X2          1   22058054   22058054  157.0596 < 2e-16 ***
X4          1     541647     541647    3.8567 0.05019 .
X5          1   79002640   79002640  562.5212 < 2e-16 ***
X6          1     285448     285448    2.0325 0.15469
X3          1     325166     325166    2.3153 0.12884
Residuals 433   60812180     140444
\end{verbatim}


\begin{verbatim}
> anova(lm(Y ~ X1 + X2 + X5 + X6 + X3 + X4, data=dat))
Analysis of Variance Table
Response: Y
           Df     Sum Sq    Mean Sq   F value    Pr(>F)
X1          1 1243181164 1243181164 8851.8032 < 2.2e-16 ***
X2          1   22058054   22058054  157.0596 < 2.2e-16 ***
X5          1   78070132   78070132  555.8815 < 2.2e-16 ***
X6          1     230467     230467    1.6410 0.2008749
X3          1     215515     215515    1.5345 0.2161052
X4          1    1638788    1638788   11.6686 0.0006956 ***
Residuals 433   60812180     140444
\end{verbatim}

\begin{verbatim}
> anova(lm(Y ~ X1 + X2 + X6 + X3 + X4 + X5, data=dat))
Analysis of Variance Table
Response: Y
           Df     Sum Sq    Mean Sq   F value    Pr(>F)
X1          1 1243181164 1243181164 8851.8032 < 2.2e-16 ***
X2          1   22058054   22058054  157.0596 < 2.2e-16 ***
X6          1    1032359    1032359    7.3507   0.00697 **
X3          1    3443417    3443417   24.5181 1.057e-06 ***
X4          1     610295     610295    4.3455   0.03769 *
X5          1   75068830   75068830  534.5114 < 2.2e-16 ***
Residuals 433   60812180     140444
\end{verbatim}


## (a)  

For each of the X3, X4, X5, and X6, calculate the coefficient of partial determination given that X1 and X2 are included in the model.

\begin{bluebox}
Let's calculate these four coefficients of partial determination:

$$
R^2_{Y 3 | 1,2} = \frac{SSR(X_3|X_1, X_2)}{SSE(X_1, X_2)} = \frac{4063370}{140967082} = 0.028825,
$$
$$
R^2_{Y 4 | 1,2} = \frac{SSR(X_3|X_1, X_2)}{SSE(X_1, X_2)} = \frac{541647}{140967082} = 0.003842,
$$

$$
R^2_{Y 5 | 1,2} = \frac{SSR(X_3|X_1, X_2)}{SSE(X_1, X_2)} = \frac{78070132}{140967082} = 0.553818,
$$

$$
R^2_{Y 6 | 1,2} = \frac{SSR(X_3|X_1, X_2)}{SSE(X_1, X_2)} = \frac{1032359}{140967082} = 0.007323,
$$
\end{bluebox}

## (b)  

On the basis of the results in part (a), which of the four additional predictor variables is best? Is the extra sum of squares associated with this variable larger than those for the other three variables?

\begin{bluebox}
Based on the results of (a), we would consider $X_5$ to be the best variable to add to a model already containing $X_1$ and $X_2$ if our objective is to explain as much of the variation in $Y$ as we can. The extra sum of squares is indeed much larger than the extra sum of squares for the other variables undergoing the same procedure. 
\end{bluebox}

## (c)  

Using the $F$ test statistic, test whether or not the variable determined to be best in part (b) is helpful in the regression model when X1 and X2 are included in the model; $\alpha=0.01$. State the alternatives, decision rule, and conclusion. Would the $F$ test statistics for the other three potential predictor variables be as large as the one here? Discuss.

```{r}
# Critical F value:
q6c_F <- qf(0.99, 3, 440)
cat("Critical F:", q6c_F)
```

\begin{bluebox}
Using the coefficients of partial determination, we deemed $X_5$ best. Let's now test whether it is helpful to include $X_5$ in a model containing $X_1$ and $X_2$:

$$
H_0: \beta_5 = 0
$$

$$
H_1: \beta_5 \neq 0
$$
\end{bluebox}

We will execute a partial F test, and decide to reject the null hypothesis if we find a p-value less than $\alpha = 0.01$ or if we find a test statistic in excess of the critical F value: $F_{0.99, 3, 440} = 3.826$. Our test statistic will be:

$$
F_{\text{test}} = \frac{(SSR_{X_5| X_1, X_2})/1}{MSE_{X_1,X_2,X_5}} =  \frac{78070132}{140444} = 555.8815
$$

The p-value associated with this test is extremely small, and we will reject the null hypothesis in favor of the conclusion that it is helpful to include $X_5$ in a model containing just $X_1$ and $X_2$. We would not find the test statistic for the other predictors to be as large as the one we found here because the numerator extra residual sum of squares introduced by the new predictor would be smaller than for $X_5$ with the denominator remaining the same, leading to a smaller test statistic.

## (d) 

Conduct an appropriate test using level of significance 0.05 to assess whether predictors $X_3,$ $X_4,$ and $X_6$ can be simultaneously dropped from an MLR model containing the six predictors ($X1-X6$). State the null and alternative hypotheses, test statistic, decision rule, and conclusion.

```{r}
q6d_crit_f <- qf(0.95, 3, 433)
cat(q6d_crit_f)
```

\begin{bluebox}
For this question, for the model $Y_i = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + \beta_3X_{i3} + \beta_4X_{i4} + \beta_5X_{i5} + \beta_6X_{i6} + \epsilon_i$, we want to know whether the predictors $X_3$, $X_4$, $X_6$ are simultaneously $0$. So, we can formulate our test as:

$$
H_0: \beta_3 = \beta_4 = \beta_6 = 0, \text{ versus }
$$

$$
H_1: \text{At least one of } \beta_3, \beta_4, \beta_6 \neq 0,
$$

Which is equivalent to a partial F test with test statistic given by:

$$
F = \frac{
(SSE_{(X_1, X_2, X_5)} - SSE_{\text{Full}})/3 
}{
SSE_{\text{Full}}/(434)
}
=
\frac{
( (60812180 + 230467 + 215515 +1638788)- 60812180)/3 
}{
60812180/434
} 
= 4.9595
$$

We will compare our test statistic to a critical value for F, this time that critical value is $F_{0.95, 3, 433} = 2.62$ if we specify a significance level of $\alpha = 0.05$ for our test. We find our test statistic to exceed the critical value. Therefore, we can reject the null hypothesis which stated $\beta_3 = \beta_4 = \beta_6 = 0$, therefore we have evidence that at least one of the corresponding predictors contributes to the model which includes $X_1, X_2, \text{ and } X_5$, so the predictors $X_3$, $X_4$, and $X_6$ cannot be simultaneously dropped from the model.
\end{bluebox}