---
title: "STAT 5605 Homework 5"
subtitle: "March 8, 2026"
author: "Jack Bienvenue"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    latex_engine: pdflatex
    keep_tex: true 
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{1em} 
fontsize: 12pt
---


```{r setup, include=FALSE}
invisible(lapply(c("ggplot2", "dplyr", "tidyr", "car", "corrplot"), 
  function(pkg) suppressPackageStartupMessages(library(pkg, 
                                                character.only = TRUE))))
```


\newpage
# Problem 1

Continuing Problem 4 from Homework 3, an analysis was performed to study the relationship between three explanatory variables, $X_1$, $X_2$, and $X_3$ and a response variable $Y$ using the following model:
$$
Y=\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 +  \epsilon,
$$
where $\epsilon \sim N(0,\sigma^2)$ independently for $n$ observations. For this dataset, recall $\mbox{MSE}=0.15306$ with 117 error degrees of freedom.  For observation 10, we have the following SAS output: 
\begin{verbatim}
      Dep Var Predicted    Std Error           Std Error  Student  Hat Diag
Obs         y     Value Mean Predict  Residual  Residual Residual         H
10     3.1780    2.5816       0.0547    0.5964     0.387    1.540    0.0195
\end{verbatim}


## (i) 
  Compute $\hat{\sigma}^2_{(10)}$, an estimate of $\sigma^2$ based on the data with the $10^{th}$ observation deleted. [Read Slide 15 of Note 6.]
  
  \begin{bluebox}
  
  We can compute $\hat{\sigma}^2_{(10)}$ externally by rearranging the expression given by Slide 15 in Note 6:
  
  $$
  (n-k-1)\text{MSE} = (n-k-2)\text{MSE}_{(i)} + \frac{e_i^2}{1-h_{ii}},
  $$
  To get an expression for $MSE_{(i)}$, which will estimate $\sigma^2_{(10)}$ here:
  
  $$
  \text{MSE}_{(i)} = \hat{\sigma}^2_{(10)} = \frac{(n-k-1)\text{MSE} - \frac{e_i^2}{1-h_{ii}}}{(n-k-2)} ,
  $$
  Now we can substitute in the provided values from the SAS output line and the other problem information. Given that the MLR model has $x$ coefficients, and $n-p = 117$ error degrees of freedom, we can identify $n$ (sample size) and $k$ (number of non-intercept regression coefficients) as $n=121$ and $k= p-1 =4 - 1 = 3$.
  
  $$
  \text{MSE}_{(i)} = \hat{\sigma}^2_{(10)} = \frac{(121-3-1)(0.15306) - \frac{0.5964^2}{1-0.0195}}{(121-3-2)} = 0.151252 ,
  $$
  
  \end{bluebox}

## (ii)  
  For observation 10, compute the \textit{externally} Studentized residual $t_{10}$.

  \begin{bluebox}
  
  We can use the other formula from Slide 15 of Note 6 in order to calculate the externally Studentized residual for the tenth observation:
  
  $$
  t_i = e_i\left[ \frac{n-k-2}{\text{SSE}(1-h_{ii}) - e^2_i}  \right]^{1/2}.
  $$
  
  We know that we have nearly all the quantities we need from the prior subquestion, however, let's make sure that we have (or can find) the \text{SSE}. The $\text{SSE}$ is not explicitly included in our given output, so we can quickly calculate it as $\text{SSE} = \text{MSE}(n-k-1) = 0.15306({121-3-1}) = 17.90802$.
  
  So, our externally Studentized 10$^{\text{th}}$ residual calculation becomes:
  
  $$
  t_i = 0.5964\left[ \frac{121-3-2}{(17.90802)(1-0.0195) - 0.5964^2}  \right]^{1/2} = 1.5487.
  $$
  
  \end{bluebox}


\newpage
# Problem 2 

A substance used in biological and medical research is shipped by airflight to users in cartons of 1,000 ampules. The data below, involving 10 shipments, were collected on the number of times the carton was transferred from one aircraft to another  over the shipment route ($X$) and the number of ampules found to be broken upon arrival ($Y$). Suppose that we fit a simple linear regression (SLR) model of $Y$ on $X$ to this data set.

\begin{center}
\begin{tabular}{rcccccccccc} \hline
$i$ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\ \hline
$X_i$: & 1 & 0 & 2 & 0 & 3 & 1 & 0 & 1 & 2 & 0 \\
$Y_i$: & 16 & 9 & 17 & 12 & 22 & 13 & 8 & 15 & 19 & 11 \\ \hline
\end{tabular}
\end{center}

```{r}
HW5PR2 <- read.table("./HW5PR2.txt", header=TRUE)
str(HW5PR2)
n <- nrow(HW5PR2)
n
PR2 <- lm(formula = Y ~ X, data=HW5PR2)
resid <- PR2$residuals
```

## (a) 

Plot the residual $e_i$ against $X_i$ to ascertain whether any departures from the SLR model are evident. What is your conclusion? 

```{r plot1, fig.cap="Question 2 Residuals versus Predictor Plot"}
plot(HW5PR2$X, resid, pch = 20, main = "Residual vs. Predictor plot", 
     xlab = "X", ylab = "residuals")
```
  
  \begin{bluebox}
  
  We plotted the residuals against the predictor in Figure \@ref(fig:plot1). In the plot, we observe that the residuals have a funnel shape, narrowing as the plot moves to the right. This may indicate that we have an issue with our homoscedasticty assumption, suggesting that variance in the errors may decrease for increasing $X$. There are no very large residuals in this set, and the residuals appear to be centered around $0$, although we may have fewer residuals close to zero than we might expect (this statement would have to be proven though). The plot does not give us reason to refute a linear relationship as we do not observe non-linear shapes in the residual plot. We should note that while there appears to be meaningful trends in the residuals in relation to the value of $X$, our sample size is very modest ($n=10$) and patterns at this sample size may be spurious, underdeveloped, and/or misleading.
  
  \end{bluebox}
  
## (b) 

Test the normality of the residuals using the Shapiro-Wilk Test and Kolmogorov-Smirnoff test, respectively. (Note that these tests are available in R; see Example 4 for relevant R code.) What do you conclude? 

```{r}
# Shapiro-Wilk Test:
shapiro.test(resid)

# Kolmogorov-Smirnoff Test:
ks.test(resid, "pnorm")
```

  \begin{bluebox}
  
  \textbf{Shapiro-Wilk Test}
  
  The Shapiro-Wilk Test is used to test the normality of residuals coming from independent observations after model fitting. It is most often used for small sample sizes.
  
  We will assume that our observations for residuals are independent (this is not necessarily true as \textit{errors should be independent}, but residuals are not necessarily independent). While the Shapiro-Wilk test is best used for continuous data, we will attempt to apply it to our problem. We will pre-specify a significance level of $\alpha = 0.05$ and use the following hypotheses:
  
  $$
  H_0: \text{The sample comes from a normally-distributed population, and}
  $$
  
  $$
  H_1: \text{The sample does not come from a normally-distributed population.}
  $$
  
  We find a p-value of 0.2632 attained by our test. This means that we do not have evidence reject the null hypothesis which stated that the sample comes from a normally-distributed population. So, from this test we will not refute our normality assumption. 
  
  \textbf{Kolmogorov-Smirnoff Test}
  
  The Kolmogorov-Smirnoff Test is another test for normality, and it is a nonparametric test which compares the empirical CDF of data from a sample to a CDF from a hypothesized distribution. The K-S test is typically used for moderate-to-large sample sizes. Our sample size of $n=10$ is decidedly more small than we would desire for using this test. We can try to run the test, making assumptions of independence of observations again. Like the Shapiro-Wilk test, it is best to have continuous response data. 
  
  Let's pre-specify a significance level of $\alpha = 0.05$, and run the test under the hypotheses:
  
  $$
  H_0: \text{The residuals come from a N(0,1) distribution, and}
  $$
  
  $$
  H_1: \text{The residuals do not come from a N(0,1) distribution. }
  $$
  
  Note that we specify a N(0,1) distribution in particular becuase the ``ks.test(resid, ``pnorm")" execution of the test assumes a N(0,1) distribution. 
  
  We find a p-value of 0.3775 for our test, which does not provide evidence to reject the null hypothesis. Therefore, we cannot refute that the residuals are standard normal distributed by the results of this question. 
  
  Note that while we failed to refute normality assumptions in this problem, it does not necessarily mean that the data is normal or that the residuals are truly N(0,1) distributed. Simply, we do not have the evidence to disprove these assumptions. Our tests were somewhat poorly employed, so we may be wary of their results. 
  \end{bluebox}
  
## (c) 

In order to determine whether or not the error variance varies with the level of $X$, conduct the following tests. State your hypotheses and conclusion using the significance level of $\alpha=0.05$. 

### (c1) 

The Breusch-Pagan (BP) test assuming that  $\log \sigma^2_i = \gamma_0 + \gamma_1 X_i$ is applicable; and

```{r}
# BP Test Implementation:
sse <- anova(PR2)[2, 2]
HW5PR2$resid <- PR2$residuals
HW5PR2$resid_sq <- HW5PR2$resid^2
diag_bp <- lm(resid_sq~X, data = HW5PR2)
# anova(diag_bp)
ssr_star <- anova(diag_bp)[1, 2]
chi_bp = (ssr_star/2)/(sse/n)^2
print(paste("test statistic:", chi_bp))
print(paste("critical chisq:", qchisq(0.95, 1)))
print(paste("test statistic p-value:", pchisq(chi_bp, 1, lower.tail = FALSE)))
```

  \begin{bluebox}
  
  The Breusch-Pagan (B-P) Test tests for heteroscedasticity in a linear regression model. Earlier, we were worried about heteroscedasticity when we looked at our residuals versus predictor values plot. 
  
  The B-P test relies on the assumption that errors are normally distributed and that observations are independent
  
  Our hypotheses are:
  
  $$
  H_0: \text{Residuals observe homoscedasticity, and }
  $$

  $$
  H_1: \text{Residuals are heteroscedastic.}
  $$
  Before running our test, we will decide upon a significance level threshold of $\alpha = 0.05$. 
  
  Upon running our test, we find that we cannot reject the null hypothesis which stated that the residuals are homoscedastic ($\text{p-value}= 0.309$). Despite what we observed in our residual plot, this test does not provide evidence that an assumption of homoscedasticity is violated.
  
  \end{bluebox}

### (c2) 

The test of Chen, C.F. (1993) assuming Var$(\epsilon_i)=\sigma^2 \exp( \lambda X_i)$.

```{r}
# C.F. Chen Test Implementation:
r_stand <- rstandard(PR2)
hii <- influence(PR2)$hat
HW5PR2$ui <- (10/8)*(1-hii)*r_stand^2
diag_ch <- lm(ui~X, data = HW5PR2)
s_chen <- anova(diag_ch)[1, 2]/2
print(paste("Test statistic:", s_chen))
print(paste("Critical chi-squared value:", qchisq(0.95, 1)))
print(paste("Corresponding p-value:", pchisq(s_chen, 1, lower.tail = FALSE)))
```

  \begin{bluebox}
  
  C.F. Chen's Test is another test for homoscedasticity. This test requires the assumption of independent, normally-distributed errors.
  
  Our hypotheses for this test are the same as for the prior question:
  
  $$
  H_0: \text{Residuals observe homoscedasticity, and }
  $$

  $$
  H_1: \text{Residuals are heteroscedastic.}
  $$
  
  Here, the results of the test are identical to the B-P test. Again we will fail to reject the null hypotheses.
  
  \end{bluebox}



\newpage
# Problem (3) 

For the data, HW3PR1.txt, given in Problem 1 of Homework 3, answer the following questions.

```{r}
HW3PR1 <- read.table("./HW3PR1.txt", header=TRUE)
str(HW3PR1)
nrow(HW3PR1) 
mod3 <- lm(formula = Y ~ X1+X2+X3, data=HW3PR1)
summary(mod3)
HW3b <- HW3PR1
HW3b$ti <- rstudent(mod3)
# externally studentized residuals
b_ct <- qt(1-(0.05/50), 45)
b_ct
HW3b[which(abs(HW3b$ti)>=b_ct),]
HW3b$ei <- mod3$residuals
HW3b$ri <- rstandard(mod3)
HW3b$hii <- influence(mod3)$hat
HW3b[which(HW3b$hii > 8/50),]
durbinWatsonTest(mod3, alternative = "positive")
dbetas <- dfbetas(mod3)
HW3f <- HW3PR1
HW3f %>% select(case) %>% mutate(CooksD = cooks.distance(mod3),dfits = dffits(mod3)) -> HW3f
bind_cols(HW3f, dbetas) -> HW3f
colnames(HW3f) <- c("case", "CooksD", "dfits","dbeta_X0", "dbeta_X1", "dbeta_X2", "dbeta_X3")
head(HW3f)
attach(HW3PR1)
pairs(Y~X1+X2+X3, pch = 20)
MP <- cor(HW3PR1[,1:4])
MP
corrplot(MP, method='color',order='alphabet', diag=FALSE)
corrplot(MP, method='circle',order='alphabet',diag=FALSE)
corrplot.mixed(MP, order='AOE')
vif(mod3)
```

## (a) 

Divide the 50 cases into two groups, placing the 25 cases with the smallest fitted values $\hat{Y}_i$ into group 1 and the other 25 cases into group 2. Conduct the Brown-Forsythe (BF) test for constancy of the error variance using $\alpha=0.025$. State the hypotheses and your conclusion.

```{r}

```

  \begin{bluebox}
  
  The Brown-Forsythe Test is a test to evaluate...
  
  The assumptions required for the test are ...
  
  The hypotheses for this test are:
  
  $$
  H_0:
  $$
  
  $$
  H_1:
  $$
  
  From our test, we find ...
  
  \end{bluebox}
  
  
## (b) 

Obtain the externally Studentized residuals and identify any outlying $Y$ observations. Use the Bonferroni outlier test procedure with $\alpha=0.05$ (i.e., using $\alpha/50$ for each observation). State the decision rule and conclusion.  


```{r}
# Bonferroni outlier test:


```
  \begin{bluebox}
  
  The Bonferroni outlier test procedures seeks to identify 
  
  In the Bonferroni outlier test procedure, we effectively split the significance level between observations ... ELABORATE on this
  
  \end{bluebox}
  
  
## (c) 

The cases are given in consecutive weeks. Prepare a time plot for of the residuals. Is there any indication that the error terms are correlated? Discuss. 

```{r plot2, fig.cap="Time Series Plot of Residuals"}

```

  \begin{bluebox}
  
  Reviewing a time series plot of residuals in the model, we find...
  
  \end{bluebox}
  
  
## (d) 

The observations are listed in time order. Assume that the MLR model is appropriate. Test whether or not positive correlation is present; using $\alpha=0.05$. State the hypotheses and conclusion. 

```{r}

```

  \begin{bluebox}
  
  Since the observations are ordered by time, we are interested in whether...
  
  We can use a ... TEST
  
  We will use hypotheses:
  
  $$
  H_0: \text{ , and}
  $$
  $$
  H_1: \text{ }
  $$
  
  \end{bluebox}
  
  
## (e) 

Obtain the diagonal element of the hat matrix. Identify any outlying $X$ observations using the rule of thumb presented in Note 6. Are any outlying $Y$ observations also outlying $X$ observations?

  \begin{bluebox}
  
  
  
  \end{bluebox}
  
## (f) 

Obtain the DFFITs, DFBETAs, and Cook's distance values. Are any cases influential  according to these measures? Discuss. 

  \begin{bluebox}
  
  
  
  \end{bluebox}
  
## (g) 

Obtain the scatter plot matrix and the correlation matrix. What do these matrices show about pairwise linear associations among the predictors? 

  \begin{bluebox}
  
  
  
  \end{bluebox}
  
## (h) 

Find the three variance inflation factors. Do they indicate that a serious multicollinearity problem exists here? 

  \begin{bluebox}
  
  
  
  \end{bluebox}


\newpage
# Problem 4

For the data $(X_1,X_2,Y)=(-1,-1,7.2)$, $(-1,0,8.1)$, $(0,0,9.8)$. Suppose that we fit a MLR model with an intercept to this data set. In this case,  $n=3$, $k=2$, and $p=3$. Compute the hat matrix ${\bf H}={\bf X} ({\bf X}' {\bf X})^{-1} {\bf X}'$.  What are $h_{ii}$, $\hat{Y}_i$, and $e_i$ for $i=1,2,3$? 

``` {r}
# Dataframe for data
q4_df <- data.frame(
  X1 = c(-1, -1, 0),
  X2 = c(-1,  0, 0),
  Y  = c(7.2, 8.1, 9.8)
)

# MLR model fit
q4_mlr_model <- lm(Y ~ X1 + X2, data = q4_df)

# build the design matrix X
X <- model.matrix(q4_mlr_model)

# find hat matrix
H <- X %*% solve(t(X) %*% X) %*% t(X)

# print hat matrix
H   

# influence matrix
h_ii <- diag(H)

# fitted values
yhat <- fitted(q4_mlr_model)
resid <- residuals(q4_mlr_model)

# give each observation its own df
obs_1 <- data.frame(h_ii = h_ii[1], yhat = yhat[1], resid = resid[1])
obs_2 <- data.frame(h_ii = h_ii[2], yhat = yhat[2], resid = resid[2])
obs_3 <- data.frame(h_ii = h_ii[3], yhat = yhat[3], resid = resid[3])

# Report
cat("Observation 1 influence:", obs_1$h_ii,
    "; fitted value:", obs_1$yhat,
    "; residual:", obs_1$resid, "\n")

cat("Observation 2 influence:", obs_2$h_ii,
    "; fitted value:", obs_2$yhat,
    "; residual:", obs_2$resid, "\n")

cat("Observation 3 influence:", obs_3$h_ii,
    "; fitted value:", obs_3$yhat,
    "; residual:", obs_3$resid, "\n")

```

  \begin{bluebox}
  
  We are fitting an MLR model with coefficients $\beta_0, \beta_1, \beta_2$ to a dataset with only three points. Therefore, our regression plane will pass through each of the points because a plane can always pass through three points in three-dimensional space. We find all observations to have an influence of $1$, and fitted values equal to their $Y$ value (leaving all residuals as $0$).
  
  \end{bluebox}


\newpage
# Problem 5

Increased arterial blood pressure in the lungs frequently leads to the development of heart failure in patients with chronic obstructive pulmonary disease (COPD). The standard method for determining  arterial lung pressure is invasive, technically difficult, and involves some risk to the patient. The radionuclide imaging technique is a noninvasive, less risky method for estimating arterial pressure in the lungs. To investigate the predictive ability of this method, a cardiologist collected data on 19 mild to moderate COPD patients. In the data set, the response variable ($Y$) is the invasive measure of systolic pulmonary arterial pressure, and the potential noninvasive predictors are  the emptying rate of blood into the pumping chamber of the heart ($X_1$) and the ejection rate of blood pumped out of the heart into the lungs ($X_2$). 

Consider the following model:
$$
Y=\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \epsilon
$$
where $\epsilon \sim N(0,\sigma^2)$.
The resulting ANOVA table is
given below.

\begin{center}
{\bf Table 1: ANOVA Table}

\begin{tabular}{lcrrrr} \hline
\multicolumn{6}{c}{Analysis of Variance} \\ \hline
Source  & DF       & Sum of Square  & Mean Square &  F Stat & Prob $>$ F \\ \hline
Model   & $\ast$   & 4961.46426     &  $\ast$     &  $\ast$    & $\ast$ \\
Error   & $\ast$    & $\ast$     &  $\ast$  &         &   \\
C Total & $\ast$   & 8087.68421     &             &         & \\ \hline
\end{tabular}
\end{center}


## (a) 

Fill in the missing values (denoted by ``$\ast$'') in Table 1. 

  \begin{bluebox}
  
  
  
  \end{bluebox}

## (b) 

The estimates of the model parameters are given in the following table.

\begin{center}
{\bf Table 2: Parameter Estimates}

\begin{tabular}{lrrrrrr} \hline
Variable & DF & Estimate & Std Error & T Stat & Prob $>$ $|\mbox{T}|$ & VIF \\ \hline
Intercept& 1  & 84.215024 & 10.25104824  &      8.215    &   0.0001 & 0 \\
 $X_1$   & 1  & -0.587215 &  $\ast$     &    $\ast$    &   0.1522 & $\ast$ \\
 $X_2$   & 1  & -0.494677 &  0.18584188  &     -2.662    &   0.0171 & $\ast$ \\ \hline
\end{tabular}
\end{center}

### (i) 

What are the "Std Error" and  "T Stat" for $X_1$?  

  \begin{bluebox}
  
  
  
  \end{bluebox}

### (ii) 

It is calculated that the Pearson correlation coefficient  between $X_1$ and $X_2$ is $r=0.65285$.  Obtain ${\rm VIF}_1$ and ${\rm VIF}_2$ for $X_1$ and $X_2$, respectively. Is there any evidence indicating that a serious multicollinearity problem exists here?  

  \begin{bluebox}
  
  
  
  \end{bluebox}
  
### (iii) 

What is the p-value for testing the null hypothesis $H_0$: $\beta_2=0$ versus one-sided alternative hypothesis $H_1$: $\beta_2 > 0$?  

  \begin{bluebox}
  
  
  
  \end{bluebox}
  
